Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-07 demystify.tex
updates

\section{I am a member of an underrepresented group in CS}\label{sec:urm}
One question that often comes up is if you should mention, e.g., in SOP, that you are a member of an underrepresented and minority (URM) group. This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

In my personal opinion, the short answer is \emph{yes}, it \emph{does} help, especially if you can articulate your experiences and how you can contribute to diversity and inclusion in CS. Most universities have a strong commitment to diversity and inclusion and want to  
\emph{increase the number of URM students} in their programs (e.g., GMU often advertises about being one of the most diverse universities in the US). Moreover, many universities and industrial organizations also have specific scholarships and fellowships for URM students (so you should look for specific scholarships for URM students).

It is understandable that some students might feel uncomfortable mentioning this in fear that they lost a chance of working with a faculty who might be biased against URM students.  However, if someone has such an issue, then you might not want to work with them anyway (and it is likely that the university would not tolerate such behavior). In my experience, it is the opposite: many profs. are actively looking for URM students to work with, and mentioning this in your application can help you stand out.

In addition to being aware of the importance of diversity and inclusion and want to contribute to it, profs. have incentives to work with URM students. For example, we often seek grants from government sponsors such as NSF, which strongly encourage broadening participation in STEM (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}).
Also, sometimes providing this information can make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain background.

In the end, this is your choice and you do what make you feel comfortable. However, just be aware that URM is often a (huge) plus, not a minus, in CS in the US. If in doubt, seek advice from your close professors and collaborators about this.

>>>>

\section{Member of an underrepresented or LBGTQ+ groups}\label{sec:urm}
A common question is whether you should mention, e.g., in SOP, that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} and LBGTQ+ group. 
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

In my personal opinion, highlighting URM or LBGTQ+ identity can \emph{indeed be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academic.  Many universities have a strong commitment to diversity and inclusion and actively 
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships specifically for URM and LBGTQ+ students, which applicants should explore if qualify.

It is understandable that some applicants might feel uncomfortable disclosing this, fearing they would lost a chance of working with a faculty with bias.  However, if someone has such an issue, then you might not want to work with them anyway (and such bias is not acceptable in academic and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek out to work with students from diverse backgrounds and view diversity as an asset to their research group}.

Moreover, promoting diversity aligns with the broadening participation in computing (BPC) goal within academia (e.g., see the \href{https://plans.bpcnet.org/GeorgeMasonUniversity_ComputerScience_DepartmentalBPCPlan.pdf}{BPC plan at GMU CS}) and government funding agencies such as NSF (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}). 
Also, sometimes providing this information can make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain background.

Ultimately, this is a personal choice and you do what make you feel comfortable. Seeking guidance from your close professors and collaborators is a good idea. However, it is important to recognize that diversity is strongly valued and celebrated within CS programs in the US, and that you should feel empowered to share your unique experiences and perspectives.
